Distortion corrected kernel density estimate on Riemannian manifolds

### 2022-10-31 paper/paper.Rmd
polished up to section 4

Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}. The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example. In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction. The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP. Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.

Figure \autoref{fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies. As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors. Comparing the left and right panel for each row, we notice that KDE with distortion corrected, has fewer misclassified observations, and therefore outperforms KDE. For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion. For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks. These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.

@@

In \autoref{tab:fivedcors}, the correlations between ranks of true densities and estimated densites from DC-KDE and KDE are presented. For ISOMAP and LLE, KDE gives higher rank correlation by less than 0.01, but both density estimators have a very high correlation of more than 0.96. However, for Laplacian Eigenmaps and UMAP where there are more distortion in the dimension reduction process, the rank correlation for KDE is close to 0 while our distortion correction methods gives a relatively high correlation of 0.87 and 0.78 respectively.

@@

Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms. 
\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions. Compared with KDE, the percentages are always higher when distortion correction methods are used. The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$. This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the $0\%$ in KDE when no distortion correction is used.

@@

For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia].
@Cheng2021-dh propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution. The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings. Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction. We use the highest density regions plots to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones. Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.

Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the procedures here. First, for each household and each of the $336$ half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method. Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of @Cheng2021-dh. The bandwidth parameter $r$ is set as 1 in Equation \@ref(eq:denestimator).

@@

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
The typical households in the top row are similar except that there are a few spikes for household $1472$ when using KDE. As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume. It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838. In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.

@@

Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde}. Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE. As for the distributions for outliers, the middle row outliers from variable bandwidth have spikes only on Tuesday and Sunday noons, while the fixed bandwidth has an increasing electricity demand across the day of the week. The bottom row outliers both have a repeated time usage pattern, but the electricity usage amount is higher with the highest median above 3kWh. These findings show the difference in finding typical and anomalous households with different kernel density estimation methods.

>>>>

Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density. The bandwidth parameter for the DC-KDE estimator is set as $r=1$ for this example. In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented. For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96. However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied. In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.

Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies. As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors. The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction. The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP. Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.

Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE. For UMAP, almost all ground truth anomalies (points outside a $99\%$ HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion. For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than $10^{-7}$) which leads to tied ranks. These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.

@@

@@

\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the $>99\%$ and $99\%$ highest density regions. Compared with KDE, the percentages are always higher when distortion correction methods are used. The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around $83\%$ while slightly lower for UMAP at around $77\%$. This could be due to the severe distortion usually induced by the UMAP algorithm.

@@

For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week [@Hyndman2018-ia]. This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions. @Cheng2021-dh propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used. We use the highest density regions plot to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.

Although full details for data processing and dimension reduction are provided in @Cheng2021-dh, we briefly describe the process here. For each household, a discrete approximation to the distribution of electricity demand at each one of the $336$ half-hourly periods of the week is found. For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm. The bandwidth parameter $r$ is set as 1 in Equation \@ref(eq:denestimator).

@@

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
The typical households in the top row are similar except that there are a few spikes for household $1472$ (selected as most typical when using KDE without distortion correction). As for the anomalies, distortion corrected KDE tends to capture instance of unusual electricity demand where usage is very low with some sudden spikes such as ID 3243 or households with a very high baseline electricity usage only across a sub-period of time such as ID 2838. In contrast, KDE without distortion correction can select households that do not exhibit anomalous behavior at all, such as household 5146 and the selection of such houesholds as anomalies may be an artefact of distortion induced by a dimension reduction algorithm.

@@

Further insights could be gained by comparing the quantile region plots of electricity demand against the time of the week for the same typical or anomalous households in \autoref{fig:electricityhdrsfixed} and \autoref{fig:electricityhdrsvkde} which are respectively selected when distortion correction is and is not applied. Again the distribution of both typical households in the top panel has shown a repeated period-of-the-week usage pattern, with higher usage during mealtime on all seven days of the week and slightly higher usage for weekends.
This repeated pattern in a week window is also obvious for the typical household ID 1321 from DC-KDE. As for the households selected when distortion correction is applied, the middle row in \autoref{fig:electricityhdrsfixed} shows a household with small spikes at noon but only on Tuesday and Sunday. The anomalies detected without distortion correction show an increasing electricity demand across the day of the week. These findings show the difference between anomalies detected with different kernel density estimation methods.

### 2022-10-31 paper/paper.tex
polished up to section 4

In \autoref{fig:tpisomapden}, we plot the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction. Data are presented on a log scale to highlight anomalies. The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region. The middle shaded region contain anomalies that are true positives in the sense of not lying in a \(90\%\) HDR. Points lying outside squares (shown as triangles) are incorrectly classified. For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used). Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection. \autoref{fig:tptsneden} shows the same plot but for t-SNE. The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified. However, the difference between KDE with and without distortion correction is stark. These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.

@@

Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density in \autoref{fig:fivedisomapden}. The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example. In \autoref{fig:fivedisomapden}, the ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction. The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP. Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than ot equal to three.

Similar to \autoref{fig:fivedmeta}, the point shapes indicate the Gaussian mixture components and the colors indicate the distance to the centre of the distribution with outliers shown in darker colors. The grey shading area highlights the \(99\%\) highest density region area and points falling into this area are the correctly classified top \(1\%\) anomalies. Darker points lying outside the shaded squares in the top left area are true anomalies but are misclassified as such. By comparing the left and right panel for each row, we could notice that KDE with distortion corrected outperforms KDE in classifying anomalies. For UMAP, almost all ground truth top \(1\%\) anomalies are not correctly identifies using KDE, while it is the other way around with the distortion corrected. It is worth noting that for DC-KDE in the left panel, there are many points having the same rank of estimated density inside the shading area, which is due to the fact that the ground truth density are almost zeros, with values less than \(10^{-7}\). And these low-density points are all captured by our distortion correction methods regardless of the dimension reduction algorithms, which again shows the robustness of DC-KDE.

@@

Furthermore, when comparing across different dimension reduction algorithms in \autoref{fig:fivedisomapden}, almost all top outliers are correctly classified in the shading areas, which again shows the robustness of the proposed distortion correction methods to the use of dimension reduction algorithms.
\autoref{tab:fourdhdrtable} further illustrutes the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions. Compared with KDE, the percentages are always higher when distortion correction methods are used. The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\). This could be due to the severe distortion usually induced by the UMAP algorithm, which can be seen from the \(0\%\) in KDE when no distortion correction is used.

@@

For the electricity consumption data of residential individuals, it would be worthwhile to explore the distribution of electricity demand rather than the raw consumption data to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}.
\textcite{Cheng2021-dh} propose two non-Euclidean distance estimators to be used in manifold learning algorithms in case of statistical manifolds with each observation as a distribution. The same Irish smart meter data is used for identifying outliers with kernel density estimation, but they fail to consider the distortion and information loss in the 2-dimensional embeddings. Therefore, we propose in this paper to apply distortion corrected KDE for anomaly detection as the downstream task of dimension reduction. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to get the 2-D embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms with the use of distortion correction. We use the highest density regions plots to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones. Therefore, we will further examine the distributions of the anomalous households to study the typical or anomalous electricity usage patterns.

Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the procedures here. First, for each household and each of the \(336\) half-hourly periods of the week, we build the empirical distributions and apply the total variation distance estimator for statistical manifold learning, which gives a 2-dimensional embedding for each dimension reduction method. Then we apply the kernel density estimator with distortion corrected for each method to the low-dimensional embedding and compare it with the KDE results in the Section 4.4 of \textcite{Cheng2021-dh}. The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.

@@

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for both density estimation methods respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
The typical households in the top row are similar except that there are a few spikes for household \(1472\) when using KDE. As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume. It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838. In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.

>>>>

In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with left panels showing results for distortion correction and the right panel showing results without distortion correction. Data are presented on a log scale to highlight anomalies. The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a \(99\%\) highest density region. The middle shaded region contain anomalies that are true positives in the sense of not lying in a \(90\%\) HDR. Points lying outside squares (shown as triangles) are incorrectly classified. For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the \(99\%\) HDR, but are not classified as such (although they are true positives if a 90\% HDR is used). Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection. \autoref{fig:tptsneden} shows the same plot but for t-SNE. The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified. However, the difference between KDE with and without distortion correction is stark. These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.

@@

Following similar steps to \autoref{twodgaussian}, we estimate the densities of the 4-D manifold and compare them with the ground truth density. The bandwidth parameter for the DC-KDE estimator is set as \(r=1\) for this example. In \autoref{tab:fivedcors}, the rank correlations between true densities and estimated densites from DC-KDE and KDE are presented. For ISOMAP and LLE, KDE without distortion correction has a slightly higher rank correlation with the true density than for KDE with distortion correction, but these differences are negligible and both density estimators have a very high correlation of more than 0.96. However, for Laplacian Eigenmaps and UMAP there more distortion is induced throughdimension reduction, the rank correlation between estimated and true densities is close to 0 when distortion correction is not applied. In the same settings, the corresponding rank correlations are relatively high at 0.87 and 0.78 for Laplacian Eigenmaps and UMAP respectively,when distortion correction is applied.

Figure \autoref{fig:fivedisomapden} is constructed in a similar fashion to \autoref{fig:tpisomapden} to show the effect of distortion correction on the detection of anomalies. As in \autoref{fig:fivedmeta}, the point shapes indicate which of the two mixture Gaussian mixture components an observation was generated from and the colors indicate the distance from the centre of the distribution with outliers shown in darker colors. The ranks of the estimated densities are shown against ranks of true densities (on the log scale) with panels on the left showing distortion corrected KDE and panels on the right showing KDE without distortion correction. The panels from top to bottom show four dimension reduction algorithms: ISOMAP, LLE, Laplacian Eigenmaps and UMAP. Note that we exclude t-SNE algorithm in this section because it is designed mainly for low-dimensional visualization purposes and is only applicable to an embedding dimension less than or equal to three.

Comparing the left and right panel for each row, we notice that KDE with distortion correction, has fewer misclassified observations, and therefore outperforms KDE. For UMAP, almost all ground truth anomalies (points outside a \(99\%\) HDR) are not correctly detected using KDE, while almost all anomalies are correctly detected after correcting for distortion. For DC-KDE in the left panel there are several triangles in a horizontal line, this arises,due to many points having an estimated density of close to zero (values less than \(10^{-7}\)) which leads to tied ranks. These low-density points are all detected as anomalies for all dimension reduction algorithms, which again shows the robustness of DC-KDE.

@@

\autoref{tab:fourdhdrtable} further illustrates the percentage of correctly classified anomalies in the \(>99\%\) and \(99\%\) highest density regions. Compared with KDE, the percentages are always higher when distortion correction methods are used. The percentages with distortion correction are very close for ISOMAP, LLE and Laplacian Eigenmaps with a value of around \(83\%\) while slightly lower for UMAP at around \(77\%\). This could be due to the severe distortion usually induced by the UMAP algorithm.

@@

For the electricity consumption data of residential individuals, it is worthwhile to explore the distribution of electricity demand rather than the raw consumption data, so as to study the usage patterns of different households or different periods or the week \autocite{Hyndman2018-ia}. This can be considered as a case of dimension reduction on a statistical manifold, that is a manifold with elements that are probability distributions. \textcite{Cheng2021-dh} propose estimators of the Total Variation Metric and Hellinger distance between distributions that can be used in a computationally practical manner for dimension reduction on statistical manifolds. Again, we use ISOMAP, LLE, Laplacian Eigenmaps, t-SNE and UMAP for dimension reduction to obtain a 2 dimensional embedding for kernel density estimation and anomaly detection. To this end, we compare the density estimates from KDE with and without distortion correction and show how robust the anomalies are to different dimension reduction algorithms, only when distortion correction is used. We use the highest density regions plot to visualize the density estimates. However, for this real data set with unknown structure, the ground truth densities are unknown and it is not possible to tell which anomalies are the true ones.

Although full details for data processing and dimension reduction are provided in \textcite{Cheng2021-dh}, we briefly describe the process here. For each household, a discrete approximation to the distribution of electricity demand at each one of the \(336\) half-hourly periods of the week is found. For any pair of households the total variation metric can found between the distributions corresponding to any half hour of the week, and summing over these gives a distance measure between the pair of households, subsequently used for dimension reduction algorithms and the Learn matric algorithm. The bandwidth parameter \(r\) is set as 1 in Equation \eqref{eq:denestimator}.

@@

\autoref{fig:electricityoutliers} shows the electricity usage data of three households for density estimation both with and without distortion correction respectively, with the top one being the most typical household with the highest density and the bottom two being the top two outliers with the lowest densities.
The typical households in the top row are similar except that there are a few spikes for household \(1472\) when using KDE. As for the anomalies, distortion corrected KDE tends to capture the unusual electricity demand when the usage is very low or high in volume. It could also capture the unusual usage pattern when there are sudden spikes in ID 3243 or very high base electricity usage for the middle time periods in ID 2838. In contrast, KDE without distortion correction is more sensitive to spikes even when the spikes happen in a certain time window in 7049, or when the usage has an obvious time-of-week pattern with a few low electricity usage periods.
